---
title: "Simulated exposures, simulated health for one city"
date: '`r format(Sys.time(), "%d %B, %Y")`'
author: Jenna R. Krall
output:
  html_document:
    toc: true
    number_sections: true
    theme: cosmo
    toc_depth: 2
bibliography: ts.bib
---




```{r, echo = F, warning=FALSE, message=FALSE}
# load libraries
library(tidyverse)
library(broom)
library(splines)
library(tsModel)


library(RColorBrewer)
library(knitr)


# set chunk options
opts_chunk$set(message=FALSE, fig.align = "center",
               warning = F,tidy = F, echo = F)

eval1 = T
evalhealth <- T
```


# Introduction

Planned simulation study:

1. Simulated exposures, simulated health for one city
2. Real exposures, simulated health for one city
3. Real exposures, simulated health for multiple cities

This is part 1 of our planned simulation study.  


```{r funcode}
#library(timeseriessim)

source("outer-sim-fns.R")
source("inner-sim-fns.R")
```



# Methods

## Summary

Simulated data:

- True time series: We simulate "true" pollutant time series data that includes a seasonal (cold vs. warm) trend, a monthly trends, and random noise
- Error time series: We add error to this pollutant time series (details below)
- Health data: We simulate "true" health data from the "true" pollutant time series

Comparing results:

- Overall pollutant comparison: We compare the true and error pollutant time series using metrics applied to the **overall time series**.
- Frequency decomposition pollutant comparison: We compare the true and error pollutant time series using metrics applied to **frequency ranges within the time series**.
- Health compare: We estimate health associations for both the true and error pollutant time series and compare the estimated health effects.

## Simulating data

### True pollutant time series

We simulated the true pollutant time series as 

\[x_t = x^{(s)} + x^{(m)} + x^(\delta)\]

where:

- $x^{(s)} =cos( \pi/ (365 / 2)  \times d)$ is a seasonal effect (warm vs. cold) and $d$ is the sequence of dates
- $x^{(m)} =cos( \pi/ (365 / 12)  \times d)$ is a monthly effect and $d$ is the sequence of dates
- $x^{(\delta)} \sim N(0, 1)$ is random error

Then $x_t$ is scaled by its standard deviation so that it has a standard deviation of 1.  The true concentration scale pollutant time series is computed as $z_t = \exp{x_t \sigma_x + \mu_x}$.

### Error pollutant time series

We consider 4 ways of introducing error.  Let $\epsilon_t \sim N(0, \sigma_\epsilon^2)$ be random error and let $x^{(k)}$ be the $k$ frequency of the true time series, $k\in \{d, m, \delta\}$:

- Classical: $x_e = x_t + \epsilon_t$
- Berkson: $x_e = (x^{(\delta)} +\epsilon_t) / (1+ \sigma_\epsilon^2) + (x_t - x^{(\delta)})$
- Amplitude change: $x_e = x^{(k)} * b + (x_t - x^{(k)})$, so that the amplitude change of $b$ only impacts frequency $k$.
- Phase change: $x_e = Lag(x^{(k)}, b) + (x_t - x^{(k)})$, so that the shift change of $b$ only impacts frequency $k$.

Then, to obtain the concentration scale error pollution,

$z_e = \exp{x_e \sigma_x + \mu_x}$

Note: The approach taken for the classical and Berkson error is derived from Goldman et al. 2011 Env Health <https://ehjournal.biomedcentral.com/articles/10.1186/1476-069X-10-61>. 




### Health data


We simulated health count data (e.g., hospitalizations) $Y\sim Poisson(\mu)$ where $\log(\mu) = 5 + \beta_1 \times z_t^*$, where 5 was chosen to obtain a reasonable baseline number of hospitalizations.  We select $\beta_1$ based on a reasonable relative risk for a 10$\mu$g/m$^3 increase in pollutant, e.g., $\beta_1 = log(RR) / IQR = log(1.01) / 10 \approx 0.01$.



## Comparing results

### Comparing pollutant time series 

We compare the true and error pollutant time series using:

- Correlation
- Log Variance Ratio ($\log(var(\text{error}) /var(\text{truth}) )$) 

We use these metrics applied to both:

1. Overall time series  
2. Separately by frequency ranges

### Comparing estimated health associations

We fitted quasi-Poisson regression models of the health data (Y) on both the true time series and the error time series.  Each model includes a natural spline with 6 degrees of freedom per year to isolate acute health effects. We compare the estimated health associations by:

- Mean bias
- Root mean squared error
- 95% confidence interval coverage


# Results


```{r setup}
# PM2.5 in NYC
mns <- 1.9
sds <- .6
sderr <- 0.47

# 0 for berkson, classical.  2/12 for amp/phas
timetrends <- c(0, 0, rep(c(2, 12), 3))
magerrs <- c(1.5, 1, 60, 14, 2, 2, .01, .01)
errs <- c("berkson", "classical", rep(c("shift", "amplitude"), each = 2), "amplitude", "amplitude")

# restrict
inds <- seq(1, length(errs))
timetrends <- timetrends[inds]
magerrs <- magerrs[inds]
errs <- errs[inds]
```


```{r setuphealth, eval = evalhealth}
#######
# health values (if needed)
rr <- c(1.001, 1.005, 1.01, 1.02, 1.05, 1.1)
rr <- 1.01
iqrt <- 10
#iqrt <- 6
beta0 <- 5
```


## Sample time series

Visualize the error/true time series under different error scenarios.



```{r samp}

inds <- seq(1, length(errs))
day1 <- NULL
check1 <- T
for(i in inds) {
  # generate pollution data
  zex <- genpoll(1000, errtype = errs[i], mean = mns[1], sd = sds[1], magerr = magerrs[i],
                timetrend = timetrends[i], check = check1, day = day1)
  
  # feed simulated day into next iteration
  if(i == 1) {
    check1 <- F
    day1 <- zex$day
    zex <- zex$zs
  }
  
  # mutate and save
  zex <- mutate(zex, timetrend = timetrends[i], errtype = errs[i], magerr = magerrs[i])
  if(i == 1) {
    zall <- zex
  } else {
    zall <- full_join(zall, zex)
  }
}

zall1 <- zall
zall <- filter(zall, date %in% seq(1, 365)) %>%
  mutate(., timetrend = ifelse(errtype %in% c("berkson", "classical"), "Effect=acute", ifelse(timetrend == 2, "Effect=seasonal", ifelse(timetrend == 12, "Effect=monthly", timetrend))),
         facet = paste0(errtype, ", MagError=", magerr, ", ", timetrend))

ggplot(zall, aes(x = date, y = value, group = type, colour = type)) +
  geom_line() +
  facet_wrap( ~ facet, ncol = 2)
```


```{r, eval = F}
# Exploration to check 12 df / year, and association for long term trends and y
i <- 1
nyears <- 10
zex <- genpoll(nyears * 365, errtype = errs[i], mean = mns[1], sd = sds[1], magerr = magerrs[i],
                timetrend = timetrends[i], check = T)

dates <- filter(zex$zs, type == "truth")$date
glm1 <- lm(zex$xt ~ ns(dates, 12 * nyears))
new <- predict(glm1, data.frame(dates))
resid <- glm1$residuals

# 12 df/year captures trends
plot(dates, zex$xt, xlim = c(1, 365))
points(dates, new, col = "red")

plot(dates, resid, xlim = c(1, 365))
  

beta1 <- log(1.01) / 10
h1 <- spread(zex$zs, type, value) %>%
  mutate(., y = healthsim(truth, beta0 = 3, beta1 = beta1))
summary(h1$y)
plot(h1$y)
```




```{r sim, eval = F}
res <- outersim(errs, mns, sds, magerrs, timetrends, nsim = 1000, N = 365 * 10,
                     health = T)
save(res, file = "simpoll-simhealth-1.RData")
```

```{r}
load("simpoll-simhealth-1.RData")
zout <- res$zout %>%
  mutate(., timetrend = factor(timetrend, levels = c("seasonal", "monthly", "acute")))
healthall <- mutate(res$healthall,timetrend = factor(timetrend, levels = c("seasonal", "monthly", "acute")))


```



## Frequency decomposition: correlation


Table of mean correlations 

```{r tabcor, eval = T}
tab1 <- getmeansd(zout) %>% filter(., meas == "cor") %>% select(., cut : mean1) %>% spread(., cut, mean1)
kable(tab1, digits = 2)
```

### Box plots of all simulations

```{r plotcor, fig.height = 6, fig.width = 8}

ggplot(zout, aes(x = cut, y = cor, group = cut, colour = cut)) +
  geom_boxplot() +
  geom_hline(yintercept = 1, linetype = 2, colour = "grey50") +
  ylab("Correlation") +
    xlab("Frequency") +  facet_wrap( error ~ timetrend , ncol = 3, scales = "free_y") +
  theme_bw() +
  theme(legend.position = "top", axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))

```


## Frequency decomposition: log variance ratio


Table of mean log variance ratios 

```{r tablvr, eval = T}
tab1 <- getmeansd(zout) %>% filter(., meas == "lvr") %>% select(., cut : mean1) %>% spread(., cut, mean1)
kable(tab1, digits = 2)
```

### Box plots of all simulations

```{r lvr, fig.height = 6, fig.width = 8}


ggplot(zout, aes(x = cut, y = lvr, group = cut, colour = cut)) +
  geom_boxplot() +
  geom_hline(yintercept = 0, linetype = 2, colour = "grey50") +

  ylab("log(var(error) / var(truth))") +
    xlab("Frequency") +  facet_wrap( error ~ timetrend , ncol = 3, scales = "free_y") +
  theme_bw() +
  theme(legend.position = "top", axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))



```



## Overall RMSE

```{r rmseall, eval = T}
ggplot(filter(zout, !is.na(rmse), cut == "overall"), aes(x = timetrend, y = rmse, group = timetrend, colour = timetrend)) +
  geom_boxplot() +
  ylab("RMSE") +
  geom_hline(yintercept = 0, linetype = 2) +
  xlab("Frequency") +
  facet_wrap(~ error, scales = "free_y") +
  theme_bw() +
  theme(legend.position = "top", axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))
```


# Health


## Mean bias (as a percent of beta)


```{r, eval = evalhealth}

out <- mutate(healthall, rmse = rmse * 1000, beta1 = log(rr) / iqrt, mbiasscale = meanbias / beta1,
              id = paste0(mean, sd, magerr, error, timetrend))
#print(out)
out1 <- select(out, type, mbiasscale, error, timetrend, id) %>%
  rename(., typeerror = error) %>%
  spread(., type, mbiasscale) %>%
  arrange(., typeerror, timetrend) %>%
  select(., -id)
kable(out1, digits = 2)
```

```{r, fig.height = 3, fig.width = 5, eval = F}
#plots
ggplot(out, aes(x = timetrend, y = meanbias, colour = type)) +
  geom_point(position = position_dodge(0.02)) +
  xlab("error timescale") +
  ylab("Mean bias") +
facet_wrap(~error)
```

## RMSE * 1000

```{r,eval = evalhealth}
out1 <- select(out, type, rmse, error, timetrend) %>%
    rename(., typeerror = error) %>%
  spread(., type, rmse) %>%
  arrange(., typeerror, timetrend)
kable(out1, digits = 2)
```


```{r, fig.height = 3, fig.width = 5, eval = F}


ggplot(out, aes(x = timetrend, y = rmse, colour = type)) +
  geom_point(position = position_dodge(0.02)) +
  xlab("error timescale") +
  ylab("RMSE") +
  facet_wrap(~error)

```
